Sperm should evolve to make female meiosis fair.

### 2014-05-18 writeup_and_notes/male_dependent_drive.tex
removing ybs

%FIgures 
{\bf Abstract:}
Genomic conflicts arise when an allele gains an evolutionary advantage at a cost to organismal fitness. 
Because only one of four products of female meiosis is transmitted to
offspring, o\"{o}genesis is inherently susceptible to such conflicts,
as alleles can compete to be transmitted into the egg. 
%(meiotic drive).  
Genomics conflicts drive the rapid evolution of suppression mechanisms, and
	many facets of  o\"{o}genesis and gametogenesis have been 
	interpreted as mechanisms of protection against genomic outlaws. 
Why then is female meiosis often left uncompleted until after fertilization - 
	potentially providing an opportunity for sperm alleles to
        meddle with its outcome in their own favor? %and facilitate meiotic drive? 
The population genetic theory presented herein suggests that
	sperm nearly always evolve to increase the fairness of female
        meiosis in the face of genomic conflicts. %%felt this
                                %%might be an important addition as
                                %%they will spread if no cost to organism
These results are consistent with current knowledge of sperm-dependent meiotic drivers 
	\yb{(loci whose distortion of female meiosis depends on sperm genotype),} 
	and suggest that the requirement of fertilization for the
        completion of female meiosis, and variation in stage of meiosis that
        fertilization occurs at may represent a mechanism employed by females 
	 to ensure a fair meiosis. %prevent meiotic drive.
\newpage

@@

We develop population genetic models exploring the evolution of sperm influence on female meiotic drive. %, focussing on models in which `self-promoting' alleles in sperm facilitate drive of like genotypes during gametogenesis in heterozygous females. 
We first focus on models in which `self-promoting' alleles in
  sperm facilitate drive of like alleles during gametogenesis in heterozygous females. 
%In this article we explore through simple population genetic models the consequences of alleles that influence the outcome of female meiosis.
These models show that such sperm-acting alleles 
	\yb{have more difficulty invading a population than do traditional meiotic drivers, 
	 and under most circumstances, cannot be maintained as a balanced polymorphism.
Because self-promoting drivers are unlikely to create a sustained genomic conflict, 
	female meiosis will have little incentive to evolve resistance to them.} 
We then examine models in which a \yb{novel} sperm-acting allele modifies the efficacy of a polymorphic meiotic driver. 
Such model universally reveal that sperm-acting drive modifiers are favored only if the suppress drive. 
These results highlight the fact that the interests of sperm and \yb{maternal}  genomes' are often aligned, as both are invested in the fate of the resultant zygote (as was speculated for the \emph{In} locus \citep{Pomiankowski1993}).
Therefore, there is little selective benefit to females in preventing sperm to influence female meioses.
Rather, features of o\"{o}genesis  may evolve to facilitate an interaction with sperm %with female meiosis
	as yet another mechanism to enforce an equitable transfer of genetic material.

@@

We first assume that the modifier is tightly
        linked to the drive locus (effectively creating a third
        allele/haplotype at this locus) and arises on the drive-background. 
%We complement this single-locus approach %of an allele with pleiotropic effects in both sexes, 
%	with alternative models of two linked loci - one a female driver, 
%	and the other, a sperm allele which modifies the effect of drive upon fertilization. 
%In this model, a female meiotic driver with no sperm-dependency 
%	initially reaches drive-viability equilibrium (with two alleles
%	$A$ and $B$ are the ancestral non-driver and driver alleles). 
%Subsequently, a sperm-acting modifier of female meiotic drive arises tightly
 %      linked to this balanced locus (effectively creating a third
 %       allele/haplotype at this locus). 
Tight linkage offers the best
        chance for a collaboration to evolve between a driver and
       an enhancer of the drive, as recombination breaks up drive haplotypes \citep{Thomson1974,Charlesworth1978,Haig1991}. 
Additionally, tight linkage between female driver and sperm modifier is consistent with the nature of well characterized drive systems which are often maintained as polymorphic inversions with numerous linked modifiers \cite{Burt2006}. 
\yb{We conclude by analyzing models with alternative linkage relationship between driver and drive modifier - 
	in Model 5 the modifier arises tightly linked to the non-driving allele, 
	and in Model 6 it is unlinked to the driver.}

@@

The interests of fertilizing sperm and female are quite well aligned during syngamy -- producing a viable and potentially fit offspring. 
%As noted above there are even opportunities for coevolution between sperm-based suppressors of female meiotic drive and the female genome.
Thus it seems plausible that aspects of oogenesis may actively evolve to facilitate the influence of sperm on ensuring a fair  outcome of female meiosis which generates fit offspring.
In light of that it is interesting to consider whether the phylogenetic variation in the stage of oogenesis when fertilization takes place reflects not
	only species' different reproductive ecologies, but also different selection pressures shaping the interaction of sperm alleles with female meiosis. 
\yb{this feels weak if we don't say anything more explicit. I also think maybe we simply state predictions:
	(1) MII drivers should drive less often when fertilized by complimentary sperm. 
	(2) Drive may be more severs in interspecific crosses in which sperm have not specifically evolved to resists  a driver. 
	(3) ??? Any more specific hypotheses -- maybe making this last sentence more explicit, e.g. self-fertilizing or otherwise frequently inbreeding animals will not require fertilization for the completion of meiosis... [note - this runs counter to our empirical examples of mice and C-elegens]}
%\gc{Add short paragraph about sex chromosomes here? XY systems don't care. ZW systems (males ZZ, females ZW). If there is
%sex ratio distortion how could this play out?}

%\yb{DOES THIS GO ANYWHERE? }

% Males can potentially influence the outcome of meiosis even before fertilization. 
% For example in Drosophila and other taxa, sperm accessory
%         gland proteins (ACPs) increase the female recombination rate
%         \cite{SOMEthING}, and there is some evidence that there is
%         variance across male genotypes in there effect on female
%         recombination rate \cite{Stevison2012}.
% Presumably this is a pleiotropic consequence of the resolution of double strand breaks initiated by the stressful/toxic composition of ACPs. 
% However, a pleiotropic effect of this is to increase the recombination rate which can decrease the effectiveness of female meiotic drivers
% active during MI \citep{Brandvain}. 
% We speculate that other effects of sperm on female meiosis may also be tolerated by females as a way of ensuring meiotic fairness -- 
% 	that is females evolve to allow male influence on female meiosis in cases in which the male influence is  egalitarian.

%Gametogenesis and meiosis are common times for genetic conflict. 
%This is a consequence of the fact that as the set of alleles that consitute
%	an individual are in the process of parting ways, 
%	with only a subset of them being assured transmission.
%In constrast, the perspective of the fertilizing sperm's genome is different, it has to live with the consequences of any distortion it causes 
%to female meiosis through the entire next life cycle of the organism. 
%It is therefore in the interest of the fertilizing sperm's genome to
%make a high fitness zygote even if that means alleles in sperm
%disrupting other copies of themselves from gaining a 
%transmission advantage through female meiosis.
  \appendix

>>>>

%FIgures 
{\bf Abstract:}
Genomic conflicts arise when an allele gains an evolutionary advantage at a cost to organismal fitness. 
Because only one of four products of female meiosis is transmitted to
offspring, o\"{o}genesis is inherently susceptible to such conflicts,
as alleles can compete to be transmitted into the egg. 
%(meiotic drive).  
Genomics conflicts drive the rapid evolution of suppression mechanisms, and
	many facets of  o\"{o}genesis and gametogenesis have been 
	interpreted as mechanisms of protection against genomic outlaws. 
Why then is female meiosis often left uncompleted until after fertilization - 
	potentially providing an opportunity for sperm alleles to
        meddle with its outcome in their own favor? %and facilitate meiotic drive? 
The population genetic theory presented herein suggests that
	sperm nearly always evolve to increase the fairness of female
        meiosis in the face of genomic conflicts. %%felt this
                                %%might be an important addition as
                                %%they will spread if no cost to organism
These results are consistent with current knowledge of sperm-dependent meiotic drivers 
(loci whose distortion of female meiosis depends on sperm genotype),}
	and suggest that the requirement of fertilization for the
        completion of female meiosis, and variation in stage of meiosis that
        fertilization occurs at may represent a mechanism employed by females 
	 to ensure a fair meiosis. %prevent meiotic drive.
\newpage

@@

We develop population genetic models exploring the evolution of sperm influence on female meiotic drive. %, focussing on models in which `self-promoting' alleles in sperm facilitate drive of like genotypes during gametogenesis in heterozygous females. 
We first focus on models in which `self-promoting' alleles in
  sperm facilitate drive of like alleles during gametogenesis in heterozygous females. 
%In this article we explore through simple population genetic models the consequences of alleles that influence the outcome of female meiosis.
These models show that such sperm-acting alleles 
	have more difficulty invading a population than do traditional meiotic drivers, 
	 and under most circumstances, cannot be maintained as a balanced polymorphism.
Because self-promoting drivers are unlikely to create a sustained genomic conflict, 
	female meiosis will have little incentive to evolve resistance to them.
We then examine models in which a novel sperm-acting allele modifies the efficacy of a polymorphic meiotic driver. 
Such model universally reveal that sperm-acting drive modifiers are favored only if the suppress drive. 
These results highlight the fact that the interests of sperm and maternal  genomes' are often aligned, as both are invested in the fate of the resultant zygote (as was speculated for the \emph{In} locus \citep{Pomiankowski1993}).
Therefore, there is little selective benefit to females in preventing sperm to influence female meioses.
Rather, features of o\"{o}genesis  may evolve to facilitate an interaction with sperm %with female meiosis
	as yet another mechanism to enforce an equitable transfer of genetic material.

@@

We first assume that the modifier is tightly
        linked to the drive locus (effectively creating a third
        allele/haplotype at this locus) and arises on the drive-background. 
%We complement this single-locus approach %of an allele with pleiotropic effects in both sexes, 
%	with alternative models of two linked loci - one a female driver, 
%	and the other, a sperm allele which modifies the effect of drive upon fertilization. 
%In this model, a female meiotic driver with no sperm-dependency 
%	initially reaches drive-viability equilibrium (with two alleles
%	$A$ and $B$ are the ancestral non-driver and driver alleles). 
%Subsequently, a sperm-acting modifier of female meiotic drive arises tightly
 %      linked to this balanced locus (effectively creating a third
 %       allele/haplotype at this locus). 
Tight linkage offers the best
        chance for a collaboration to evolve between a driver and
       an enhancer of the drive, as recombination breaks up drive haplotypes \citep{Thomson1974,Charlesworth1978,Haig1991}. 
Additionally, tight linkage between female driver and sperm modifier is consistent with the nature of well characterized drive systems which are often maintained as polymorphic inversions with numerous linked modifiers \cite{Burt2006}. 
We conclude by analyzing models with alternative linkage relationship between driver and drive modifier - 
	in Model 5 the modifier arises tightly linked to the non-driving allele, 
	and in Model 6 it is unlinked to the driver.

@@

The interests of fertilizing sperm and female are quite well aligned during syngamy -- producing a viable and potentially fit offspring. 
%As noted above there are even opportunities for coevolution between sperm-based suppressors of female meiotic drive and the female genome.
Thus it seems plausible that aspects of oogenesis may actively evolve to facilitate the influence of sperm on ensuring a fair  outcome of female meiosis which generates fit offspring.
In light of that it is interesting to consider whether the phylogenetic variation in the stage of oogenesis when fertilization takes place reflects not
	only species' different reproductive ecologies, but also different selection pressures shaping the interaction of sperm alleles with female meiosis. 
We predict that female drive should be

\yb{this feels weak if we don't say anything more explicit. I also think maybe we simply state predictions:
	(1) MII drivers should drive less often when fertilized by complimentary sperm. 
	(2) Drive may be more severs in interspecific crosses in which sperm have not specifically evolved to resists a driver. 
	(3) ??? Any more specific hypotheses -- maybe making this last sentence more explicit, e.g. self-fertilizing or otherwise frequently inbreeding animals will not require fertilization for the completion of meiosis... [note - this runs counter to our empirical examples of mice and C-elegens]}

  \appendix
